Statistical Theory\\ Chess Dataset Analysis

### 2024-09-30 Article/Article.tex
update bibliography, link to github and pdf

\begin{abstract}
This study explores the relationship between \textbf{player ratings} and \textbf{chess game duration} using a dataset of online games. We aim to determine whether \textbf{higher-rated players} tend to play longer or shorter games and investigate other factors influencing game length, such as \textbf{time controls} and \textbf{game outcomes}. A cleaned dataset was analyzed through statistical techniques, including the \textbf{Kolmogorov-Smirnov test} for normality, the \textbf{Mann-Whitney U test} for differences in game length between rating groups, and \textbf{correlation analysis}. Results indicate that \textbf{higher-rated players} generally engage in longer games, while greater disparities in player ratings lead to shorter matches. Additionally, \textbf{draws} tend to prolong game duration compared to decisive outcomes. The game's turn count distribution shows positive skewness and excess kurtosis, leading us to model it with a \textbf{Gamma-Poisson mixture}. This model effectively captures the underlying distribution characteristics, providing further insights into the statistical patterns governing chess match dynamics.
\end{abstract}

@@

\subsection{Rating Distribution}
The distribution of player ratings in the dataset initially resembles a normal distribution with a peak around the mean. However, this peak diminishes when games involving unrated players are excluded, bringing the distribution closer to normality. This is illustrated in Figure~\ref{fig:transformation_plot}. Additionally, the QQ plot (Figure~\ref{fig:rating_qq}) supports the similarity between the player rating distribution and a normal distribution, albeit with deviations at the tails, which are exacerbated by the QQ plot's sensitivity to deviations at the tail.

>>>>

\begin{abstract}
This study explores the relationship between \textbf{player ratings} and \textbf{chess game duration} using a dataset of online games. We aim to determine whether \textbf{higher-rated players} tend to play longer or shorter games and investigate other factors influencing game length, such as \textbf{time controls} and \textbf{game outcomes}. A cleaned dataset was analyzed through statistical techniques, including the \textbf{Kolmogorov-Smirnov test} for normality, the \textbf{Mann-Whitney U test} for differences in game length between rating groups, and \textbf{correlation analysis}. Results indicate that \textbf{higher-rated players} generally engage in longer games, while greater disparities in player ratings lead to shorter matches. Additionally, \textbf{draws} tend to prolong game duration compared to decisive outcomes. The game's turn count distribution shows positive skewness and excess kurtosis, leading us to model it with a \textbf{Gamma-Poisson mixture}. This model effectively captures the underlying distribution characteristics, providing further insights into the statistical patterns governing chess match dynamics. \newline \newline
The code for this project is available at \href{https://github.com/nakardn/Chess-Statistical-Analysis}{https://github.com/nakardn/Chess-Statistical-Analysis}
\end{abstract}

@@

\subsection{Rating Distribution}
The distribution of player ratings in the dataset initially resembles a normal distribution with a peak around the mean. However, this peak diminishes when games involving unrated players are excluded, bringing the distribution closer to normality. This is illustrated in Figure~\ref{fig:transformation_plot}. Additionally, the QQ plot (Figure~\ref{fig:rating_qq}) supports the similarity between the player rating distribution and a normal distribution, albeit with deviations at the tails, which are exacerbated by the QQ plot's sensitivity to deviations at the tail \cite{qqplotwiki}.
